Semantic Adversarial Autoencoder for Zero-Shot Learning

### 2017-12-13 doc/paper.tex
Revised the paper.

\title{Adversarial Zero-Shot Learning}

@@

\begin{abstract}
In Zero-shot learning(ZSL), how to learn discriminative features for fine-grained classes is a main challenge. In this paper, we propose a novel model: semantic adversarial autoencoder(SAAE) to solve ZSL problem. This approach is based on a semantic autoencoder(SAE), which projects image representation onto the space of semantic features. We incorporate a discriminator for semantic representations to  strengthen the discriminative power of the model. The result of experiments on two standard datasets shows that our model is able to outperform other baseline models, especially in generalized ZSL scenario. 
\end{abstract}

@@

Zero-shot learning has been an active research topic in the field of image classification. It simulates how people learn things: when people see an unfamiliar object, they use existing knowledge and try to evaluate the new object. Ideally, a large-scale image classification system should be able to recognize novel categories based on its previous training experience. One of the main challenges for object recognition is the lack of sufficient annotations for all possible concepts. This problem becomes even more severe when we target at the task of fine-grained classification because the annotation is more expensive and the number of fine-grained classes is huge. Realizing this limitation, researchers resort to additional information, for example, textual description, to solve this problem.

On the other hand, autoencoder has shown its great express power to present complicate distributions such human faces, natural sceneries and natural language. It is able to convert complicated real world data distributions, e.g. text features, to very well-formed low dimension. Many studies have shown that after applying autoencoder techniques, the learned latent feature space often has clear semantic meanings. For example, in \cite{makhzani2015adversarial}, all the digits can be well embedded into a low-dimensional manifold such that similar digits will have smaller distance within that manifold.

 \par

The fundamental problem of ZSL is to extract semantically meaningful feature embeddings that could bridge the gap between similar image/text features and fine-grained classes. Autoencoder becomes an ideal option for this task because it can automatically learn important features. In \cite{kodirov2017semantic}, they present a novel solution to zero-shot learning--Semantic AutoEncoder(SAE). Taking the encoder-decoder paradigm, an encoder aims to project a visual feature vector into the semantic space as in the existing zero-learning shot models and the decoder should be able to reconstruct the original visual feature. Their results show that under this framework, the learned projection function from the seen classes is able to generalize better to the new unseen classes. One shortcoming of this model is that the encoder tends to learn similar features across multiple classes, which is disappointing for the task of fine-grained classification.

Inspired by these models, we propose an Semantic Adversarial AutoEncoder(SAAE). In our model, we have an encoder that projects a visual feature vector into the semantic space and a decoder that decodes and reconstructs the original visual features. Besides, we incorporates a discriminator as in \cite{makhzani2015adversarial} to ensure that generating from any part of prior space results in meaningful samples.

Two types of discriminators will be discussed in this project, one is independent from semantic space and the other is dependent from semantic space.

@@

\section{Related Works}    
Considering that fine-grained categories might share some common knowledge, a common approach is to seek for an intermediate semantic representation that could connect seen and unseen classes. Human specified attributes are first explored to represent the discriminative properties shared among both seen and unseen categories in zero-shot learning\cite{torresani2010efficient, lampert2009learning}. One limitation of this method is that the creation of attributes still relies on human labors, making it difficult to scale up to meet large scale needs. \par

@@

Apart from handcrafting attributes, another scheme is to directly use the online textual document as the additional information source. \cite{Elhoseiny2013} is one of the first works to use Wiki documents as text attributes. \cite{Ba_2015_ICCV} proposed a model that changes both the ways of extracting features from images and text domains. More specifically, the image features are extracted from the activation layer of a convolutional neural network(CNN), and then go through a linear projection layer to reduce dimensions. The input text, e.g. Wikipedia articles, is first converted to one-hot encodings of the words with their tf-idf scores, which can be viewed as attributes, and then fed into a multi-layer perceptron (MLP) to generate a deep representation, with the same dimension as the final image feature, and unique for each class. The final prediction is obtained by the dot product of the two generated features. Instead of learning an embedding space for each modalities, \cite{frome2013devise} learns joint image-word embeddings so as to embedding images and sentences into a common space.\\

@@

@@

	There are two scenarios for ZSL in image classification, namely standard ZSL and generalized ZSL. During training time, given training images $\bb{x}^c$ and their corresponding semantic representations $\bb{t}^c$ for $N_s$ seen classes $c$, the model learns to bridge the gap between $x^c$ and $t^c$.

	During testing time, for standard ZSL, we are given images $\bb{x}^{uc}$ from $N_{us}$ unseen classes and semantic representations $\bb{t}^{uc}$ for those unseen classes without knowing the correspondence between $\bb{x}^{uc}$ and $\bb{t}^{uc}$. The task is to determine which unseen class each image belongs to. For generalized ZSL, the test images are not constrained to unseen classes: images from both seen and unseen classes can be used as test images and semantic representations for both classes should be provided.

During testing time, we follow the same classification method as in \cite{kodirov2017semantic}. The input test image $X$ is first projected using weight matrix $W$ to get the semantic representation $H$ and then the test class is decided by
\begin{align}
	\hat{c} = \textrm{arg} \min_c \textrm{dist}(\bb{t}^c, \bb{h}),
\end{align}
where dist is the distance function, which can be the negative cosine similarity or the Euclidean distance, We find that cosine similarity in general gives higher classification accuracy than the Euclidean distance does. We use the former in our experiments.

@@

\subsection{SAAE-exp}
Our extension with adding adversarial part for the autoencoder does not improve the accuracy for the strict ZSL scenario. We experiment with a lot of other settings, such as adding more layers for autoencoder, changing the hidden dimensions of generator and discriminator for adversarial part, and different sampling strategies, such as sampling from hidden state to produce negative samples, as described in \cite{makhzani2015adversarial} section 2. None of them give any performance boost.

We suspect the reason could be two-folded. The first reason is that the semantic features plays the essential role to the model. We also try ablation study that does not include the semantic space as constraints. In this case the model is just slightly better than random guess. This is expected since ZSL task usually deals with fine-grained image classification, and without the additional semantic space input, the models trained only based on image features are likely to have a poor result. However, the only way we figure out to incorporate the semantic features into the adversarial autoencoder is to minimize the l2 loss between the hidden state of the encoder and the semantic representation, and this major constraint bound the accuracy of the model.

The second possible reason is that the prior is hard to pick for the semantic distribution. The prior serves to regularize the encoded samples so that the hidden state of an encoder is similar to the prior distribution we choose. In normal cases, for example alignDraw\cite{mansimov2015generating}, gaussian prior works fine. However, in out scenario, the output from encoder has to be close to the semantic space, thus our chosen prior has to resemble the distribution of the semantic space. As discussed before, the semantic space is the probability that one attribute appears in this class. If we view the attributes in the space of all possible classes, this is similar to a multinomial distribution. However, changing gaussian prior to multinomial distribution also does not work well.

The SAAE-exp model performs the best in generalized classification task. We think this is as expected due to the nature of our task. Adversarial autoencoder is proposed to regularize the class distribution with a shape. This regularized shape is even more important in fine-grained classification task we are dealing with, where the difference of classes is very subtle. The regularized shape can give a better split between different classes, and this will help improve the classification accuracy in generalized ZSL scenario.

@@

@@

In this paper, we proposed to use the adversarial antoencoder framework incorporating semantic features as a solution to ZSL problem. Our method named semantic adversarial autoencoder(SAAE) projects the image representation into the semantic feature space with a discriminator matching the projection to a given prior. It achieves comparable accuracies on AWA and CUB datasets, especially in the generalized ZSL scenario. We also demonstrate that this new SAAE model is more robust than existing SAE model through extensive experiments.

>>>>

\title{Semantic Adversarial Autoencoder for Zero-Shot Learning}

@@

\begin{abstract}

In zero-shot learning tasks, how to classify the data in the unseen classes in one domain by leveraging the information from the semantic domain is a main challenge. We propose a novel model, the semantic adversarial autoencoder, to address this challenge. By injecting the global information of the distribution of the semantic representation, the model acquires better generalization ability. We perform experiments on the AwA and CUB datasets, and show that our model outperforms the semantic autoencoder in the generalized zero-shot learning tasks.

\end{abstract}

@@

Zero-shot learning (ZSL) has been an active research topic in the field of image classification. It simulates how people learn things: when people see an unfamiliar object, they use existing knowledge and try to evaluate the new object. Ideally, a large-scale image classification system should be able to recognize novel categories based on its previous training experience. One of the main challenges for object recognition is the lack of sufficient annotations for all possible concepts. This problem becomes even more severe when we target at the task of fine-grained classification because the annotation is more expensive and the number of fine-grained classes is huge. Realizing this limitation, researchers resort to additional information, for example, textual or attributive description, to solve this problem.

On the other hand, autoencoder has shown its great express power to present complicate distributions such human faces, natural sceneries, and natural language. It is able to convert complicated real world data distributions, e.g. text features, to very well-formed low dimension. Many studies have shown that after applying autoencoder techniques, the learned latent feature space often has clear semantic meanings. For example, in \cite{makhzani2015adversarial}, all the digits can be well embedded into a low-dimensional manifold such that similar digits will have smaller distance within that manifold.

The fundamental problem of ZSL is to extract semantically meaningful feature embeddings that could bridge the gap between similar image/text features and fine-grained classes. Autoencoder becomes an ideal option for this task because it can automatically learn representative features. In \cite{kodirov2017semantic}, the authors present a novel solution to zero-shot learning, the Semantic Autoencoder (SAE). Taking the encoder-decoder paradigm, an encoder aims to project a visual feature vector into the semantic space as in the existing zero-learning shot models, and the decoder should be able to reconstruct the original visual feature. Their results show that under this framework, the learned projection function from the seen classes is able to generalize better to the new unseen classes. One shortcoming of this model is that the encoder tends to learn similar features across multiple classes, which is disappointing for the task of fine-grained classification.

Inspired by these models, we propose the model of Semantic adversarial autoencoder (SAAE). In this model, we have an encoder that projects a visual feature vector into the semantic space and a decoder that decodes and reconstructs the original visual features. Besides, we incorporates a discriminator as in \cite{makhzani2015adversarial} to ensure that generating from any part of prior space results in meaningful samples. Two types of models that use different priors are investigated. In one type of the model, the prior is the Gaussian noise prior, and in the other type the semantic representation itself is used as the prior.

@@

\section{Related Works}    
Considering that fine-grained categories might share some common knowledge, a common approach is to seek for an intermediate semantic representation that could connect seen and unseen classes. Human specified attributes are first explored to represent the discriminative properties shared among both seen and unseen categories in zero-shot learning \cite{torresani2010efficient, lampert2009learning}. One limitation of this method is that the creation of attributes still relies on human labors, making it difficult to scale up to meet large scale needs. \par

@@

Apart from handcrafting attributes, another scheme is to directly use the online textual document as the additional information source. \cite{Elhoseiny2013} is one of the first works to use Wiki documents as text attributes. \cite{Ba_2015_ICCV} proposed a model that changes both the ways of extracting features from images and text domains. More specifically, the image features are extracted from the activation layer of a convolutional neural network (CNN), and then go through a linear projection layer to reduce dimensions. The input text, e.g. Wikipedia articles, is first converted to one-hot encodings of the words with their tf-idf scores, which can be viewed as attributes, and then fed into a multi-layer perceptron (MLP) to generate a deep representation, with the same dimension as the final image feature, and unique for each class. The final prediction is obtained by the dot product of the two generated features. Instead of learning an embedding space for each modalities, \cite{frome2013devise} learns joint image-word embeddings so as to embedding images and sentences into a common space.\\

@@

	Note that the SAE is more like a domain transfer machine, rather than a standard autoencoder: It is not minimizing the reconstruction loss; it is minimizing the two directions of transfers between the two domains. Later in our experiments we find that in practice the learning may be majorly driven by one of the two directions (see section \ref{subsec:SAE}).

@@

	There are two scenarios for ZSL, namely standard ZSL and generalized ZSL. For the standard ZSL, During training time, given training images $\bb{x}^c$ and their corresponding semantic representations $\bb{t}^c$ for $N_s$ seen classes $c$, the model learns to bridge the gap between $x^c$ and $t^c$. During testing time, we are given images $\bb{x}^{uc}$ from $N_{us}$ unseen classes and semantic representations $\bb{t}^{uc}$ for those unseen classes without knowing the correspondence between $\bb{x}^{uc}$ and $\bb{t}^{uc}$. The task is to determine which unseen class each image belongs to. For the generalized ZSL, the training procedure is the same, but the test images are not constrained to unseen classes: images from both seen and unseen classes can be used as test images and semantic representations for both classes are provided.

During the test time, we follow the same classification method as in \cite{kodirov2017semantic}. The input test image $\bb{x}$ is first projected to the semantic representation $\bb{h}$, and the class is predicted by
\begin{align}
	\hat{c} = \textrm{arg} \min_c \textrm{dist}(\bb{t}^c, \bb{h}),
\end{align}
where dist is the distance function. The distance function can be the negative cosine similarity, $\frac{(\bb{t}^c)^{\top} \bb{h}}{||\bb{t}^c||_2 ||\bb{h}||_2}$, or the Euclidean distance, $||\bb{t}^c - \bb{h}||_2$. We find that cosine similarity in general gives higher classification accuracy than the Euclidean distance does. We use the former in our experiments.

@@

\subsection{SAAE-exp}

Our extension with adding adversarial part for the autoencoder does not yield higher accuracy than that achieved by the SAE in the standard ZSL scenario. We have experimented with a lot of other settings, such as adding more layers for the autoencoder, changing the hidden dimensions of the generator and the discriminator in the adversarial part, and different sampling strategies (such as sampling from the hidden state to produce negative samples, as described in \cite{makhzani2015adversarial}, section 2), but none of them gives any performance boost.

We suspect the reason is that the semantic features plays the essential role in the standard ZSL tasks. We also try ablation study that does not include the semantic space as constraints. In this case the model is just slightly better than random guess. This is expected since ZSL task usually deals with fine-grained image classification, and without the additional semantic space input, the models trained only based on image features are likely to have a poor result. However, the only way we figure out to incorporate the semantic features into the adversarial autoencoder is to minimize the L2 loss between the hidden state of the encoder and the semantic representation, and this major constraint limits the accuracy that can be achieved by the model.

% The second possible reason is that the prior is hard to pick for the semantic distribution. The prior serves to regularize the encoded samples so that the hidden state of an encoder is similar to the prior distribution we choose. In normal cases, for example alignDraw\cite{mansimov2015generating}, gaussian prior works fine. However, in out scenario, the output from encoder has to be close to the semantic space, thus our chosen prior has to resemble the distribution of the semantic space. As discussed before, the semantic space is the probability that one attribute appears in this class. If we view the attributes in the space of all possible classes, this is similar to a multinomial distribution. However, changing gaussian prior to multinomial distribution also does not work well.

The SAAE-exp model performs the best in generalized classification task. We think this is as expected due to the nature of the task. Adversarial autoencoder is proposed to inject the global information of the distribution of the semantic representation of the training data. Through testing the model on both the seen and unseen classes, the model performs well because it has learned the distribution over all the seen classes.

@@

\subsection{SAE}
\label{subsec:SAE}

We note that SAE gives the best performance when the coefficient $\lambda$ in \eqref{eq:SAELoss} is much smaller than 1. This indicates that the learning is actually driven by learning how to generate (from attribute to reconstructed image), rather than how to infer (from image to attribute). We think this is an observation that has not be pointed out in the original SAE paper \cite{kodirov2017semantic}.

@@

In this paper, we proposed to use the adversarial antoencoder framework incorporating semantic features as a solution to ZSL problem. Our method, the semantic adversarial autoencoder, projects the image representation into the semantic feature space with a discriminator matching the projection to a given prior. It outperforms the semantic autoencoder in the generalized zero-shot learning tasks on the AwA and CUB datasets.
